---
title: "SAD with a seed bank"
author: "Jay T. Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
   - \usepackage{graphics}
output:
  pdf_document:
  fig_caption: true
---

# Initial Setup
```{r results='hide', message=FALSE}
rm(list=ls())
setwd("~/GitHub/seedSAD/")

# Import Scripts
source("bin/DiversityFunctions.r")
```

# Background

Data are from experiment where Rpf was added to soils containing Brassica.
After applying Rpf for 6 weeks, we sequenced RNA and DNA 16S.
Kuo et al. (2020)

# Load sequence data

```{r}
# Define inputs
shared <- "data/Rpf/mothur/output/Brassica.bac.final.shared"
design.in <- "data/Rpf/data/design.txt"
level  <-  "0.03"

# Import shared file
OTU <- read.otu(shared, "0.03")

# Remove OTUs with less than two occurrences across all sites 
OTU <- OTU[, which(colSums(OTU) >= 2)]

# Remove mock community 
OTU <- OTU[1:20, ]  
```

# Load experimental design

```{r}
# Import design
design <- read.delim(design.in, header = T, row.names = 1)

# Match OTU matrix and design
design_red <- design[which(rownames(design) %in% rownames(OTU)),]

# Order by Rpf treatment then sample pairing
redesign <- design_red[order(design_red$treatment, design_red$pair),]

# Sort OTU matrix by treatment orderings
OTU.match <- OTU[match(row.names(redesign), row.names(OTU)),]
```

# Create SADs

```{r}
# Raw abundance OTU matrix
OTU <- OTU.match

# Presence-absence OTU matrix
dataPA <- (OTU > 0)*1

# Relative abundance OTU matrix
dataREL <- OTU
  for(i in 1:nrow(OTU)){
  dataREL[i,] = OTU[i,]/sum(OTU[i,])
}

# Function for ranking OTUs
RAC <- function(x = ""){
  x = as.vector(x)
  x.ab = x[x > 0]
  x.ab.ranked = x.ab[order(x.ab, decreasing = TRUE)]
  return(x.ab.ranked)
}

# Apply RAC function to relative abundance matrix
ranks <- apply(dataREL, 1, RAC)

# Convert list of OTU ranks (with different lengths) to a data frame
rank.df <- sapply(ranks, "length<-", max(lengths(ranks))) 

# Create single rank vector for all samples 
max.rank <- lengths(ranks)
rank.vector<- seq(1, max(max.rank),1)

# # Parse out samples using grep
# rank.DNA <- rank.df[,grep("BrassicaD", colnames(rank.df), value = TRUE)]
# rank.DNA <- data.frame(rank.vector, rank.DNA)
# 
# rank.RNA <- rank.df[,grep("BrassicaC", colnames(rank.df), value = TRUE)]
# rank.RNA <- data.frame(rank.vector, rank.RNA)

# DNA sequences for -Rpf (ctr) and +Rpf (rpf)
rank.ctr <- data.frame(rank.vector, rank.df[,6:10])
rank.rpf <- data.frame(rank.vector, rank.df[,16:20])
```

# Plot SADs

Plotting all SADs for DNA, where samples are paired by +Rpf and -Rpf treatment

```{r}
# Initiate file export
# pdf("figures/Fig1.AllSADs.pdf", width = 5, height = 5)
# 
# # Set up basic plotting dimensions
# opar <- par(no.readonly = TRUE)  # Saves plot defaults
# par(opar)
# par(mar = c(4.5, 5, 1, 1), oma = c(0.5, 0.5, 0.5, 0.5))
# 
# # Create plot 
# plot(rank.DNA[,1], rank.DNA[,2],
#      xlim = c(0, 8000), ylim = c(0.000001, 0.1), log = "y", type ="n",
#      las = 1, bg = rgb(0, 0, 1, 0.5), 
#      axes = T, xlab = "", ylab = "")
# 
# mtext(expression('Rank'), side = 1,
#       outer = FALSE, cex = 1.25, line = 4, adj = 0.5)
# 
# mtext(expression('Relative Abundance'), side = 2,
#       outer = FALSE, cex = 1.25, line = 4, adj = 0.5)
# 
# box(lwd = 1.5)
# 
# # Add lines for DNA-based SADs (color = red)
# for(i in 2:dim(rank.DNA)[2]){ # number of lines to draw 
#   points(rank.DNA[,1], rank.DNA[,i ], 
#          col = rgb(1, 0, 0, 0.5), cex = 0.2, type = "l", lwd = 1.5)
# }
# 
# # Add lines for DNA-based SADs (color = blue)
# for(i in 2:dim(rank.RNA)[2]){ # number of lines to draw 
#   points(rank.RNA[,1], rank.RNA[,i ], 
#          col = rgb(0, 0, 1, 0.5), cex = 0.2, type = "l", lwd = 1.5)
# }
# 
# legend(3500, 0.085, legend=c("DNA", "RNA"), box.lty = 0,
#        col=c("red", "blue"), lty = 1, cex = 1.0)
# 
# #Terminate plot 
# dev.off()
```

# Plot pairs of SADs in multipanel figure

```{r}
# Initiate file export

pdf("~/GitHub/seedSAD/figures/SAD_Rpf_paired.pdf", width = 5, height = 5)

# Multiplot orientation
par(mfrow = c(2, 2)) # four plots per page 

for (i in 2:ncol(rank.ctr)){
  plot(rank.ctr[,1], rank.ctr[,i],
         col = "red", cex = 0.2, type = "l", lwd = 1.5, las = 1,
         xlim = c(0, 8000), ylim = c(0.000001, 0.1), log = "y",
         axes = T, xlab = "Rank", ylab = "Relative Abundance", 
         main = colnames(rank.ctr[i]))

  legend(5000, 0.095, legend=c("-Rpf", "+Rpf"), box.lty = 0,
       col=c("red", "blue"), lty = 1, cex = 0.8)

  points(rank.rpf[,1], rank.rpf[,i], col = "blue", cex = 0.2, 
         type = "l", lwd = 1.5)
}     

# Close plot device
dev.off()
graphics.off()
```

# Make Preston-like SADs

```{r}
# Using `sads` package to cacluate octave frequencies
require("sads")

# Split up OTU matrix by DNA and RNA
otu.cont <- OTU[grep("D1|D2|D3|D4|D5", rownames(OTU), value = TRUE),]
otu.rpf <- OTU[grep("DR1|DR2|DR3|DR4|DR5", rownames(OTU), value = TRUE),]

# Calculate octave frequency for abundance classes 
oct.cont <- apply(otu.cont, 1, octav)
oct.rpf <- apply(otu.rpf, 1, octav)

# Initiate file export
pdf("figures/SAD_Rpf_dist.pdf", width = 5, height = 5)

# Multiplot orientation
par(mfrow = c(2, 2)) # four plots per page 

for (i in 1:length(oct.cont)){
  plot(unlist(oct.cont[[i]][1] - 0.25), unlist(oct.cont[[i]][3]),
         col = "red", cex = 0.2, type = "h", lwd = 3, las = 1,
         xlim = c(0, 15), ylim = c(0, 3000),
         axes = T, xlab = "log2(Abundance class)", ylab = "N of species",
         main = names(oct.cont[i]))
  
  legend(6, 2000, legend = c("-Rpf", "+Rpf"), box.lty = 0,
       col=c("red", "blue"), lty = 1, cex = 1.0)
        
  points(unlist(oct.rpf[[i]][1] + 0.25), unlist(oct.rpf[[i]][3]), col = "blue", 
         cex = 0.2, type = "h", lwd = 4)
}
         
# Close plot device
dev.off()
graphics.off()
```