---
title: "SAD with a seed bank"
author: "Jay T. Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
   - \usepackage{graphics}
output:
  pdf_document:
  fig_caption: true
---

# Initial Setup
```{r results='hide', message=FALSE}
rm(list=ls())
setwd("~/GitHub/seedSAD/")

# Save Default Plot Settings
#opar <- par(no.readonly = TRUE)  # Saves plot defaults

# Import Required Packages
library("png")
library("grid")
#library("vegan")
#library("labdsv")
#library("car")

# Import Scripts
source("bin/DiversityFunctions.r")
```

# Background

Data are from Evolution Canyon, Israel (Muscrella, Lennon et al. in prep). 
Evolution canyon has a north-facing, mesic slope and a south-facing, xeric slope.
On each slope, replicate soil samples were taken.
We extracted DNA and RNA from each soil sample and sequenced 16S rRNA. 
Therefore, there is a pair of RNA and DNA sequences for each sample. 
Here, we construct species abundance curves for each sample. 
Goal: determine if RNA-based SAD is different from DNA-based SAD. 
What is the best SAD to fit the data? (lognormal, Zipf, logseries, broken stick)?
Are parameters for a given model different when comparing RNA and DNA. 
Xiaotian suggegsts looking into Gambin model.
Also, Wasserstein distance between each SAD pair applied in deep learning.

# Load data

```{r}
# Define inputs
shared <- "data/EC.bac.final.shared"
design.in <- "data/design.txt"
level  <-  "0.03"

# Import shared file
ec_data <- read.otu(shared, "0.03")

# Note: owing to amplification issues, we only sequenced 76 of 80 samples.
# The four samples not included are C-1E-R, EC-2G-R, EC-2J-R, EC-6I-D
dim(ec_data)
counts <- count.groups(otu.matrix = ec_data)
low.coverage <- counts[which(counts < 50000)]

# Remove low coverage samples 
ec_data2 <- ec_data[-c(which(rownames(ec_data) %in% c(names(low.coverage)))), ]

# Remove problematic samples (EC_2A_D, EC_2A_R, EC_2C_R, EC_2D_R)
#ec_data.tmp <- ec_data2[-c(9,20:21,24:27,32,37,74),]
ec_data3 <- ec_data2[-c(which(rownames(ec_data2) %in% 
                     c("EC_2A_D", "EC_2A_R","EC_2C_R", "EC_2D_R"))), ]

# Remove pairs of problematic samples
ec_data4 <- ec_data3[-c(which(rownames(ec_data3) %in% 
                     c("EC_2A_R", "EC_2A_D","EC_2C_D", "EC_2D_D"))), ]

# Remove OTUs with less than two observations across samples
ec_data_red <- ec_data4[,which(colSums(ec_data4) >= 5)]

# Rarefied
floor(min(rowSums(ec_data_red)))
```

# Load experiment design information

Again, at Evolution Canyon, there are two slopes: 
The xeric, south-facing slope is referred to as the African Slope (AF)
The mesic, north-facing slope is referred to as the European Slope (EU)

On each of the slopes, there are two stations:
On AF, there are stations 1 and 2.
On EU, there are stations 5 and 6. 
At each of these stations, there is transect containing 10 sites. 

Reminder, at each site, RNA and DNA were sequenced from a single soil sample. 

```{r}
# Import design
design <- read.delim(design.in, header = T, row.names = 1)

# Match OTU matrix and design
design_red <- design[which(rownames(design) %in% rownames(ec_data_red)),]

# Create factors for model
slope <- design_red$slope # factor 1
molecule <- design_red$molecule # factor 2
paired <- design_red$paired_across_slope
site <- design_red$site
station <- design_red$station
slope.molecule <- data.frame(cbind(as.character(slope),
                  as.character(molecule))) # Y matrix with factor 1 and 2
slope.molecule.concat <- do.call(paste, 
                                 c(slope.molecule[c("X1", "X2")],
                                   sep = "")) # create unique treat ID vector

# Create a vector of molecules by station
station.molecule.concat <- paste(station, molecule, sep = "")
pair.station <- c(rep(seq(1:9), each=2), rep(seq(1:5), each=2), 
                  rep(seq(1:10), each=2), rep(seq(1:9), each=2))
```

# Create SADs

```{r}
# Raw abundance OTU matrix
otu.matrix <- ec_data_red

# Presence-absence OTU matrix
dataPA <- (otu.matrix > 0)*1

# Relative abundance OTU matrix
dataREL <- otu.matrix
  for(i in 1:nrow(otu.matrix)){
  dataREL[i,] = otu.matrix[i,]/sum(otu.matrix[i,])
}

# Log-transform relative abundance matrix
#dataREL.log <- decostand(dataREL,method="log")

# Function for ranking OTUs
RAC <- function(x = ""){
  x = as.vector(x)
  x.ab = x[x > 0]
  x.ab.ranked = x.ab[order(x.ab, decreasing = TRUE)]
  return(x.ab.ranked)
}

# Apply RA function to relative abundance matrix
ranks <- apply(dataREL, 1, RAC)

# Convert list of OTU ranks (with different lengths) to a data frame
rank.df <- sapply(ranks, "length<-", max(lengths(ranks))) 

# Create single rank vector for all samples 
max.rank <- lengths(ranks)
rank.vector<- seq(1, max(max.rank),1)

# Parse out DNA and RNA samples
rank.DNA <- rank.df[,grep("_D", colnames(rank.df), value = TRUE)]
rank.DNA <- data.frame(rank.vector, rank.DNA)

rank.RNA <- rank.df[,grep("_R", colnames(rank.df), value = TRUE)]
rank.RNA <- data.frame(rank.vector, rank.RNA)
```

# Plot SADs

In this section, just plotting all DNA- and RNA-based SADs. 
Does not take into account station, pairing, etc. 

```{r}
# Initiate file export
pdf("figures/Fig1.AllSADs.pdf", width = 5, height = 5)

# Set up basic plotting dimensions
opar <- par(no.readonly = TRUE)  # Saves plot defaults
par(opar)
par(mar = c(4.5, 5, 1, 1), oma = c(0.5, 0.5, 0.5, 0.5))

# Create plot 
plot(rank.DNA[,1], rank.DNA[,2],
     xlim = c(0, 8000), ylim = c(0.000001, 0.1), log = "y", type ="n",
     las = 1, bg = rgb(0, 0, 1, 0.5), 
     axes = T, xlab = "", ylab = "")

mtext(expression('Rank'), side = 1,
      outer = FALSE, cex = 1.25, line = 4, adj = 0.5)

mtext(expression('Relative Abundance'), side = 2,
      outer = FALSE, cex = 1.25, line = 4, adj = 0.5)

box(lwd = 1.5)

# Add lines for DNA-based SADs (color = red)
for(i in 2:dim(rank.DNA)[2]){ # number of lines to draw 
  points(rank.DNA[,1], rank.DNA[,i ], 
         col = rgb(1, 0, 0, 0.5), cex = 0.2, type = "l", lwd = 1.5)
}

# Add lines for DNA-based SADs (color = blue)
for(i in 2:dim(rank.RNA)[2]){ # number of lines to draw 
  points(rank.RNA[,1], rank.RNA[,i ], 
         col = rgb(0, 0, 1, 0.5), cex = 0.2, type = "l", lwd = 1.5)
}

legend(3500, 0.085, legend=c("DNA", "RNA"), box.lty = 0,
       col=c("red", "blue"), lty = 1, cex = 1.0)

#Terminate plot 
dev.off()
```

# Plot pairs of DNA- and RNA-based SADs

```{r}
# Initiate file export
pdf("figures/Fig2.multiSADpaired.pdf", width = 5, height = 5)

# Multiplot orientation
par(mfrow = c(2, 2)) # four plots per page 

for (i in 2:ncol(rank.DNA)){
  plot(rank.DNA[,1], rank.DNA[,i],
         col = "red", cex = 0.2, type = "l", lwd = 1.5, las = 1,
         xlim = c(0, 8000), ylim = c(0.000001, 0.1), log = "y",
         axes = T, xlab = "Rank", ylab = "Relative Abundance")
  
  legend(2500, 0.085, legend=c("DNA", "RNA"), box.lty = 0,
       col=c("red", "blue"), lty=1, cex = 1.0)
        
  points(rank.RNA[,1], rank.RNA[,j], col = "blue", cex = 0.2, 
         type = "l", lwd = 1.5)
}
         
dev.off()
```